---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(nycflights13)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)
```

## Tutorial 1: the basics

This tutorial has three parts, all based on a different chapter of R for data science (https://r4ds.hadley.nz/).

- The basics (chapter 2)
- data transformation (chapter 3)
- data import (chapter 7)


## The basics

This first part of the tutorial walks you through the bases of programming in R, it roughly follows the second chapter of R for data science.


### R is a calculator 

Write the R code required to add fifty-nine plus seventy-three plus two, divided by three:

```{r basicOperations, exercise=TRUE}

```

```{r basicOperations-solution}
(59 + 73 + 2) / 3
```

```{r basicOperations-check}
grade_this({
  pass_if_equal()

  fail_if_equal(
    59 + 73 + 2 / 3,
    message = "Make sure to pay attention to the order of operations. Use parentheses () where necessary."
  )
  
  fail()
})
```


### Creating objects

You can use the `<-` (assignment operator) to create an object. You can then use code to manipulate objects.
Assign the value 2 to an object called two:

```{r objects, exercise=TRUE}

two

```

```{r objects-hint}
#Example
one <- 1

one # Print the value of the `two` object
```

```{r objects-solution}
# with implicit printing ----
two <- 2
two # Print the value of the `two` object

# with explicit printing ----
two <- 2
print(two) # Print the value of the `two` object

# without printing ----
two <- 2
```

```{r objects-check}

grade_this_code()

```

In most programming languages, the assignment is done with the `=` operator. If you use `=` in R, it will work (most of the time), but it is not the standard way, so make sure you use `<-`.


### the `c()` function

The `c()` function allows you to create vectors containing several elements.
You can do operations on the resulting vector. Try it:

```{r c, exercise=TRUE}
#Example
number <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)

number

#Try an operation here:

```


### Comments

Code is not always straightforward and there are plenty of time (what I mean here is: ALL THE TIME) when an additional comment might make it easier for the reader of your code to understand what you were trying to achieve.

```{r comment, exercise=TRUE}
#This is a comment, they all start with #

# Write an operation and a comment explaining what it does:

```

```{r comment-solution}
# Example

two <- 2 # This is a comment, it explains that we are assigning the value 2 to the object two
```

### Functions

Functions are pre-written code to perform specific operations. Many are built-in R, and many more can be access through installing additional packages. 
Functions is how we do most things in R. 
For example, you can use the `sum()` function to add a large number of numbers, for example, numbers that you have stored in a vector


```{r functions, exercise=TRUE}
# This is a vector of 10,000 numbers between 1 and 100
numb <- sample(1:100, 10000, replace = TRUE)

# Use the sum() function to add all the numbers in the vector

```

```{r functions-hint-1}
# you can start by examining what we mean by a 'vector' of numbers

numb
```

```{r functions-hint-2}
# Now, pass `numb` to the `sum()` function to add all the numbers in the vector

sum()
```

```{r functions-solution}
numb <- sample(1:100, 10000, replace = TRUE)

sum(numb)
```

```{r functions-check}
grade_this_code()
```

You could then do further modifications on the results, for example divide by the number of items in your vector to get the average (you can do this with the `length()` function). Note that there is a built-in function for the average: `mean()`

## Data transformation

This part of the tutorial is roughly based on chapter 3 of R for data science. 
For a lot of the day to day data wrangling, we will use the package `dplyr`.
Like in chapter 3 of R for data science, I use the `nycflights13` package to access data on flights from New York in 2013.

First, let's look at the data.

```{r flights}
library(nycflights13)
flights


```

`dplyr` provides tools to manipulate data frame and perform many of the common data wrangling tasks that will be performed during data analysis.

We will start with operations on rows, then move on to operations on columns.

### Filter: Flights on the first of January

The code below filters the dataframe for flights on the first of January. 
Can you add code to check the number of flights that left New York on the 1st of January 2013?

```{r firstjan, exercise=TRUE}
filter(flights, day == 1 & month == 1)


```

The first argument is always a dataframe, and the output is also a dataframe.

### Distinct: Counting unique occurences

The `disctinct()` function will help find all the unique rows. 
It is specifically useful to find unique combinations of variables, for example, below, use distinct to find all the unique pairs of airports in the datasets (use the variables `origin` and `dest`):

```{r occ, exercise=TRUE}
distinct(flights, ____, ____)


```



```{r quiz}
quiz(
  question("How many unique pairs of airports are there in this dataset?",
    answer("25"),
    answer("214"),
    answer("224", correct = TRUE),
    answer("236")
  )
)
```

### Count: how many flights to a specific destination

If you want to calculate how many flights there were between two destinations, you can use the `count()` function.
Try it below. Once you have done that, you can also use an additional argument in the function, `sort`, in order to sort the result. Sort takes only the value TRUE or FALSE.

```{r count, exercise=TRUE}
count(flights, ____, ____)


```

We now move on to functions that allow you to do operations on columns.

### Mutate: the workhorse of data wrangling

The function `mutate()` allows you to compute new variables for your dataset.
For example, in the `flights` data, you can compute the speed of the aircraft during a specific flight by using the `distance` and the `air_time` variables, keep in mind that air_time is in minutes.

```{r mut, exercise=TRUE}
mutate(flights, speed = )

```

```{r mut-hint}
mutate(flights, speed = ____ / ____)

```

### Select: retain only the columns you need

The function `select` allows you to only retain the columns you need for the computation at hand.

Use the `flights` dataset to select the columns: year, month, and day. 

```{r select, exercise=TRUE}
select(flights, ____)
  # YOUR CODE HERE

```

```{r select-hint}
select(flights, ____)

# pass year, month, and day as arguments

```


Select all columns between year and day (inclusive), you can do this using the `:` operator.

```{r select2, exercise=TRUE}
select(flight, ____)

```

```{r select2-hint}
# Use `year:day` inside `select()` to select the range.

```

### Rename a column

Rename the column 'tailnum' to 'tail_num'. 
Use `select()` with a new name for the column on the left-hand side of `=`.
```{r renam, exercise=TRUE}
flights |> 
  # YOUR CODE HERE
```

```{r renam-hint}
# Use `select(new_name = old_name)` to rename columns while selecting.

```



### relocate()

Move the 'time_hour' column to the beginning of the dataset.

```{r reloc, exercise=TRUE}
flights |> 
  # YOUR CODE HERE
```

```{r reloc-hint}
# Use `relocate()` and specify the variable name.

```


### The pipe: make your code easier to write and read

The pipe (`|>`) allows you to chain operations and make complex operations easier to parse compare to a series of nested functions.

Use the pipe to filter the `flights` dataset for flights with destination "IAH".
Then group the data by `month` and summarize to find the average delay.

```{r pipe, exercise=TRUE}

```

```{r pipe-hint-1}
#Use `|>` to chain commands and `group_by()` before `summarize()
```

```{r pipe-hint-2}
flights |> 
  filter(dest == "IAH") |> 
  group_by(month) |> 
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE))

```


### group_by()

Group the `flights` dataset by `dest` and use `slice_max()` to find the most delayed flight for each destination.

Use `group_by()` and `slice_max()` to select the row with the highest value in `arr_delay`.

```{r groupby, exercise=TRUE}
flights |> 
  group_by(____) |> 
  slice_max(____)


```

```{r groupby-hint, exercise=TRUE}
# Use `group_by(dest)` first, then `slice_max(arr_delay, n = 1)` to find the highest delay per group.


```


  
## Importing data

This part of the tutorial heavily relies on [chapter 7 of R for data science](https://r4ds.hadley.nz/data-import).

Reading data from files (and writing data to files) is possibly the most necessary skill in order to be able to do any analysis. 
While it is fairly easy in practice, it requires some understanding of how and where a computer stores files and how to access those files.

In this part of the tutorial, we will load data that is contained within the virtual environment that the tutorial runs in. It will help illustrate how the filesystem and filepath work.

### Filepath

The filepath is the path to a file on your computer. It is the address that your computer uses to find the file you are looking for. To navigate the filesystem, you need to know where you are and where you want to go and how to move from one place to the other.

Files are stored in folders (also called directories). The folders are organized in a tree structure. The root of the tree is the root folder, and it contains all the other folders. The current folder (called the 'current working directory') is where you are currently working. It is denoted with a dot `.`. You can move up one folder by using `..`. 

For this tutorial, the file structure is:

```
root ('.')
|--- 01-Basics.Rmd  # This file
|--- data
|    |--- bsa2021_healthcare_equalities_recoded.csv
|    |--- bsa2021_healthcare_equalities_recoded.tab
|    |--- bsa2021_healthcare_equalities_recoded.xlsx
|    |--- bsa2021_healthcare_equalities_recoded.rds
```

We join the folder name with the file name using a forward slash `/`. So, to access the file `bsa2021_healthcare_equalities_recoded.csv`, you need to use the path `./data/bsa2021_healthcare_equalities_recoded.csv`.

There are many types of files. In this tutorial we will see how to open some of the most common data format: csv, tab separated, and excel files. We will also look at one of R data format.

### Loading a csv file

First, there is a csv (comma-separated value) file that you will be able to read using the `read.csv()` function named `bsa2021_healthcare_equalities_recoded.csv`.

```{r readcsv, exercise=TRUE}

bsa <- ____

head(bsa)
```

```{r readcsv-hint}
# The function to use is read.csv
# Then you can use the path to the file inside this function

bsa <- read.csv(____)
```

```{r readcsv-solution}

bsa <- read.csv("./data/bsa2021_healthcare_equalities_recoded.csv")

head(bsa)

```

```{r readcsv-check}
grade_this_code()
```

The data is from the British Social Attitudes survey and is a recoded version of the open teaching dataset for the 2021 edition of this survey. We will reuse this data during the class.

### Loading a tab separated file

Tab separated files are very common. Like CSV files they are simple text files. But instead of a comma, they use tabulation.
In the same data folder, there is a file named bsa2021_healthcare_equalities_recoded.tab. Load it.

```{r readtab, exercise=TRUE}

bsa <- ____

head(bsa)
```

```{r readtab-hint}
# the function to use is read.delim
#Then you can use the path to the file inside this function

```

```{r readtab-solution}

bsa <- read.delim("./data/bsa2021_healthcare_equalities_recoded.tab")

head(bsa)

```

```{r readtab-check}
grade_this_code()
```

Note that tabulation (\t) is the standard delimiter in read delim, but if you have files using a different separator, a space for example, you can use the `sep` argument to change the separator that is being used. For example, `read.delim("yourfilepath", sep=" ")` will read a file that uses a space as a separator.
Word to the wise, it is usually not a good idea to be creative with these separators, unless you have a good reason, stick with csv and tab files.
One example of good reason is if you store data in languages that use commas as a decimal marker instead of periods (French is an example of such a language). In cases like these, you can use 'read.csv2()' which expects a semicolon as a separator.

### Loading an excel file

Excel files are a bit more complicated than text files. While there are ways to load excel files, for these to work, your excel spreadsheet needs to be reasonably well formatted (i.e. it is a series of observation in rows with different variables in columns).
We can use the `readxl` package to load functions that will allow us to read excel files.

```{r readxl, exercise=TRUE}
library(readxl)
bsa <- ____

head(bsa)
```

```{r readxl-hint}
# the function to use is read_xlsx()
#Then you can use the path to the file inside this function

```

```{r readxl-solution}

bsa <- read_xlsx("./data/bsa2021_healthcare_equalities_recoded.xlsx")

head(bsa)

```

```{r readxl-check}
grade_this_code()
```

### Loading an RDS file

R also has its own format. Two of these are quite useful (RData and RDS), but for now, we will only focus on RDS.

You can use `readRDS()` to load the data that is in the data folder. 

```{r readrds, exercise=TRUE}

bsa <- ____

head(bsa)
```

```{r readrds-hint}
# the function to use is readRDS
#Then you can use the path to the file inside this function

```

```{r readrds-solution}

bsa <- readRDS("./data/bsa2021_healthcare_equalities_recoded.rds")

head(bsa)

```

```{r readrds-check}
grade_this_code()
```
