---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Tutorial 1: The basics

This tutorial has three parts, all based on a different chapter of R for data science (https://r4ds.hadley.nz/).

- The bases (chapter 2)
- data transformation (chapter 3)
- data import (chapter 7)


## The bases

This first part of the tutorial walks you through the bases of programming in R, it roughly follows the second chapter of R for data science.


### R is a calculator 

Write the R code required to add fifty-nine plus seventy-three plus two, divided by three:

```{r basicOperations, exercise=TRUE}

```

### Creating objects

You can use the `<-` (assignment operator) to create an object. You can then use code to manipulate objects.
Assign the value 2 to an object called two:

```{r objects, exercise=TRUE}
#Example
one <- 1

one

two

```


### the `c()` function

The `c()` function allows you to create vectors containing several elements.
You can do operations on the resulting vector. Try it:

```{r c, exercise=TRUE}
#Example
number <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)

number

#Try an operation here:

```


### Comments

Code is not always straightforward and there are plenty of time (what I mean here is: ALL THE TIME) when an additional comment might make it easier for the reader of your code to understand what you were trying to achieve.

```{r comment, exercise=TRUE}
#This is a comment, they all start with #

# Write an operation and a comment explaining what it does:

```


## Data transformation

Data transformation (chapter 3 of R4DS) dplyr on rows and column, the pipe, groups


## Importing data

data import (chapter 7)

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

