---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(nycflights13)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)
```

## Tutorial 1: The basics

This tutorial has three parts, all based on a different chapter of R for data science (https://r4ds.hadley.nz/).

- The bases (chapter 2)
- data transformation (chapter 3)
- data import (chapter 7)


## The bases

This first part of the tutorial walks you through the bases of programming in R, it roughly follows the second chapter of R for data science.


### R is a calculator 

Write the R code required to add fifty-nine plus seventy-three plus two, divided by three:

```{r basicOperations, exercise=TRUE}

```

### Creating objects

You can use the `<-` (assignment operator) to create an object. You can then use code to manipulate objects.
Assign the value 2 to an object called two:

```{r objects, exercise=TRUE}
#Example
one <- 1

one

two

```


### the `c()` function

The `c()` function allows you to create vectors containing several elements.
You can do operations on the resulting vector. Try it:

```{r c, exercise=TRUE}
#Example
number <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)

number

#Try an operation here:

```


### Comments

Code is not always straightforward and there are plenty of time (what I mean here is: ALL THE TIME) when an additional comment might make it easier for the reader of your code to understand what you were trying to achieve.

```{r comment, exercise=TRUE}
#This is a comment, they all start with #

# Write an operation and a comment explaining what it does:

```


### Functions

Functions are pre-written code to perform specific operations. Many are built-in R, and many more can be access through installing additional packages. 
Functions is how we do most things in R. 
For example, you can use the `sum()` function to add a large number of numbers, for example, numbers that you have stored in a vector


```{r functions, exercise=TRUE}
# This is a vector of 10,000 numbers between 1 and 100
numb <- sample(1:100, 10000, replace = TRUE)


```

You could then do further modifications on the results, for example divide by the number of items in your vector to get the average (you can do this with the `length()` function). Note that there is a built-in function for the average: `mean()`

## Data transformation

This part of the tutorial is roughly based on chapter 3 of R for data science. 
For a lot of the day to day data wrangling, we will use the package `dplyr`.
Like in chapter 3 of R for data science, I use the `nycflights13` package to access data on flights from New York in 2013.

First, let's look at the data.

```{r flights}
library(nycflights13)
flights


```

`dplyr` provides tools to manipulate data frame and perform many of the common data wrangling tasks that will be performed during data analysis.

We will start with operations on rows, then move on to operations on columns.

### Flights on the first of January

The code below filters the dataframe for flights on the first of January. 
Can you add code to check the number of flights that left New York on the 1st of January 2013?

```{r firstjan, exercise=TRUE}
filter(flights, day == 1 & month == 1)


```

## Importing data

data import (chapter 7)

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

