---
title: "Data structures and types"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(stringr)
library(forcats)

knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

In the previous tutorials, we jump right into working with data and glossed over a lot of the details. In this tutorial, we take a step back to consider what happens when we load data in memory: how does R represents that data internally and what that means for what we can do with it. We will also talk about how we can change the way R represents a specific piece of data so we can make what we want with it, be it a plot, a table, or a model.

This tutorial draws heavily on [R for data science](https://r4ds.hadley.nz/transform) chapter 12 to 18. I highly recommend you refer to these chapters for more detailed explanation of any of the topics covered below.

Let's jump in and talk about data structures. In order to manipulate data, programming languages use a number of data structures. They do not all have the same data structures and this can make it awkward when you code in more than one language. In R, the main data structures are vectors, lists, matrices, and data frames. We will cover each of these in turn. 
While we talk about data structures, we will also talk about types, types describe how data is stored in memory and what operations can be performed on it. In R, the main types are logical, integer, double, character, and factor. We will cover these at the same time we cover data structure. A useful reference for an in-depth presentation of both is the [Advanced R book](https://adv-r.hadley.nz/) (specifically chapter 3 Vectors).
Before we jump in, let's just note that R does not have a data structure for scalars, a single number, in R a scalar is just a vector of length 1.

## Vectors

A vector is a uni-dimensional data structure. It can contain elements of the same type. The most common types of vectors are logical, numeric, and character vectors.

```{r, echo=TRUE}
v <- c(1:5)
v

is.vector(v)
```


### Logical vectors

A logical vector is a vector that can only contains three values: TRUE, FALSE, and NA (missing value).

```{r, echo=TRUE}
v <- c(TRUE, FALSE, TRUE, NA)
v

is.vector(v)
```

Very often logical vectors are used as a result of some operation on data and can be very useful to subset data.

In the example below, use mutate to create the auth variable which checks whether a respondent scored above 3 on the libertarian to authoritarian scale (libauth variable).

```{r logvec, exercise=TRUE}
bsa <- readRDS("data/bsa2021_healthcare_equalities_recoded.rds")

bsa <- ____ |> ____(auth = ____ >= ____)

summary(bsa$auth)

```

```{r logvec-solution}
bsa <- readRDS("data/bsa2021_healthcare_equalities_recoded.rds")

bsa <- bsa |> mutate(auth = libauth >= 3)

summary(bsa$auth)

```

One of the great things you can do with logical vectors is use them to do Boolean algebra using the `&` (and), `|` (or), and `!` (not) operators.

Another operator that is useful and generate logical vectors as a result is the `%in%` operator. This operator checks whether the elements of the first vector are in the second vector. Use `%in%` in the code below to see how it works:

```{r in, exercise=TRUE}
1:12 ____ c(1, 3, 5, 7, 9, 11)


```


### Thing you should know about vectors

Vectors can only contain elements of the same type. If you try to create a vector with elements of different types, R will coerce them to the same type. So if you try to create a vector with a number and a character, R will coerce the number to a character.
Use the `str()` function to check the type of the elements in the vector v below:

```{r vecchar, exercise=TRUE}
v <- c(1, "a", 2, 4)



```


## Numbers

R does not have scalars so uses vector of length 1 instead to represent single numbers.
Use the `is.vector()`, `mode()`, and `typeof()` functions to check the type of the variable v below:

```{r scalar, exercise=TRUE}
v <- 1

```



## Matrices

Matrices are two dimensional arrays.
Type the name of the matrix and run the code to see it.

```{r mat, exercise=TRUE}
matrix <- matrix(data = c(1:6), nrow = 2, byrow = TRUE)

____

```

### Types

Matrix and vectors can only hold data of one type.
If you try to create a vector or matrix with data of different type, it will be of the type that allows to represent all elements of the matrix.
Look at the matrices below to see what type they are. You can type their name to see how they are displayed, and also use the function `typeof()` to see their type.

```{r mattype, exercise=TRUE}
matrix1 <- matrix(data = c(1:6), nrow = 2, byrow = TRUE)
matrix2 <- matrix(data = c(1, 2, 3, 4, 5, 3.14), nrow = 2, 
                  byrow = TRUE)
matrix3 <- matrix(data = c(1:5, "Marmalade"), nrow = 2, byrow = TRUE)

____

```


## Data frame

Data frames are two dimensional arrays (like matrices) but they can hold various types of data.
Display the data frame below to see what it looks like.

```{r df, exercise=TRUE}
dt <- data.frame(name = c("Julia", "Simon", "Christina"), 
                 age = c(24, 26, 19), 
                 sex = c("Female", "Male", "Female"))

____

```

```{r df-solution}
dt <- data.frame(name = c("Julia", "Simon", "Christina"), 
                 age = c(24, 26, 19), 
                 sex = c("Female", "Male", "Female"))

dt

```

Within each column, every element has to have the same type.

There are two important function that are useful with data frames (and other types):
`summary` and `str`, use them below.

```{r df2, exercise=TRUE, exercise.setup="df-solution"}

____(dt)

____(dt)

```

Internally, data frames are lists of vectors. This means that you can access the columns of a data frame using the `$` operator or the `[[` operator. Try accessing the age column of the dt data frame below:

```{r df3, exercise=TRUE, exercise.setup="df-solution"}

____$____

____[["____"]]

```

## Lists

Lists are very useful because of their flexibility, but this also makes them sometimes difficult to work with.

Elements of a lists can be of any type (even list). Data frames are also lists under the hood, but with limitations on their structure (no two column can have the same name, the length of every column needs to be the same).
Run the code to create the list below, then play around with the elements to display only some of them. You can access the elements of a list in two ways: using `[[]]` and `[]`, which do not produce the same result. Play around with these as well. I suggest you use the `str()` and `typeof()` functions on the output of your subsetting to see how the results differ.

```{r list, exercise=TRUE}
l <- list(c(1:3), "Today", matrix(1:6, nrow = 2))

l____
l____
```


We have now covered the basics data structures that you are likely to encounter in the course of doing data analysis. We will now switch to looking at some of the data types that R uses to represent data in memory.


## Strings

In this section of the tutorial, you will learn the basics of working with strings in R, including creating strings, using escape characters, and working with raw strings.

### Strings in R
Strings in R are text enclosed in single (`'`) or double (`"`) quotes.

```{r, echo=TRUE}
string1 <- "Hello, R!"
string2 <- 'Let\'s learn! This string contains two "quirks"!'
```

Notice how you can use escape characters (like `\'`) to include special symbols within your strings. We also use single quotes to enclose the string when it contains double quotes and vice versa.

Fill in the blanks to create valid strings:

```{r string, exercise = TRUE}
# Create strings using single and double quotes
string1 <- ___
string2 <- ___

```


### What Are Escape Characters?
Escape characters allow you to include special characters in strings. Common examples include:

- `\` for backslash
- `"` for double quotes
- `\'` for single quotes
- `\n` for a new line
- `\t` for a tab

```{r, echo=TRUE}
escaped_string <- "He said, \"Learn R!\""

escaped_string
```

### Practice with Escape Characters

Write a string that spans two lines and includes a quote:

```{r escapechar, exercise = TRUE}
# Create a string with a new line and a quote
multi_line_string <- ___

multi_line_string
```



### Raw Strings

Raw strings (`r""`) are useful when you want to avoid interpreting escape sequences, such as in file paths or regular expressions.

```{r, echo=TRUE}
raw_string <- r"(C:\Users\Name)"

raw_string
```

### Create a Raw String

Write a raw string for a file path:

```{r rawstring, exercise = TRUE}
# Create a raw string for a file path (it does not need to be an actual path, you can make up one in this instance)
my_path <- ___

my_path
```


### Quiz

QUESTIONS NOT GREAT - NEED BETTER ONES

```{r, echo=TRUE, eval=FALSE}
print("This is a \"quote\".")
```

```{r quiz1}
quiz(
  question("What is the output of the code above?",
    answer("Error"),
    answer('This is a \"quote\".', correct = TRUE),
    answer("This is a 'quote'.")
  ),
  question("Which syntax correctly creates a string containing a file path?",
    answer(r"("C:\\\\Users\\\\Documents")", correct = TRUE),
    answer(r"("C:\\Users\\Documents")"),
    answer(r"(r\"(C:\\Users\\Documents)\")")
  )
)
```


### Combining Strings

The `str_c()` function from the `stringr` package is used to combine multiple strings.

```{r, echo=TRUE}
library(stringr)
combined <- str_c("Hello", "world", sep = ", ")
print(combined)
```

### Combine Strings
Combine the following words into a single string, separated by a space:

```{r comb_s, exercise = TRUE}
# Combine strings
word1 <- "Learning"
word2 <- "R"
word3 <- "is"
word4 <- "fun!"
combined <- str_c(___, ___, ___, ___, sep = " ")

combined
```


### Handling Missing Values
By default, `str_c()` does not replace `NA` with "NA" when combining strings, and missing values are contagious.

```{r, echo=TRUE}
with_na <- str_c("Hello", NA, "world")
without_na <- str_c(str_replace_na(c("Hello", NA, "world")))

with_na

without_na
```

### Manage Missing Values
Combine the following strings and deal with `NA` values:

```{r missingvalue, exercise = TRUE}
# Manage missing values
part1 <- "Learning"
part2 <- NA
part3 <- "R"
result <- str_c(part1, part2, part3)

result
```




```{r quiz2}
quiz(
  question("What does the `sep` argument in `str_c()` do?",
    answer("Replaces missing values."),
    answer("Controls the separator between strings.", correct = TRUE),
    answer("Collapses multiple strings into one.")
  ),
  question("What is the output of the following code? `str_c(\"Data\", \"Science\", sep = \"-\")`",
    answer("Data-Science", correct = TRUE),
    answer("DataScience"),
    answer("NA")
  )
)
```


### Extracting and Modifying Strings

Use `str_sub()` to extract parts of a string by specifying the start and end positions.

```{r, echo=TRUE}
# Extract characters 1 to 5
substring <- str_sub("Learning R is fun!", 1, 5)

substring
```

### Extract Substrings
Extract the word "Learning" from the string:

```{r substring, exercise = TRUE}
text <- "Learning R is fun!"
word <- str_sub(text, ___, ___)

word
```


### Replacing Substrings
Use `str_replace()` or `str_replace_all()` to replace parts of a string.

```{r, echo=TRUE}
# Replace "fun" with "awesome"
replaced <- str_replace("Learning R is fun!", "fun", "awesome")

replaced
```


Replace "R" with "data science" in the following string:

```{r replacetext, exercise = TRUE}
text <- "Learning R is fun!"
updated_text <- str_replace(text, ___, ___)

updated_text

```

### Splitting Strings

Use `str_split()` to divide a string into parts based on a pattern.

```{r, echo=TRUE}
# Split a sentence into words
words <- str_split("Learning R is fun!", " ")

words
```

Split the following string into individual words:

```{r splitstring, exercise = TRUE}
text <- "Learning R is fun!"
words <- str_split(text, ___)

words
```

### Joining Strings

Use `str_flatten()` to join a list of strings into one string.

```{r, echo=TRUE}
# Join words into a sentence
sentence <- str_flatten(c("Learning", "R", "is", "fun!"), " ")

sentence
```

Join the following words into a sentence:

```{r flatten, exercise = TRUE}
words <- c("Learning", "R", "is", "fun!")
sentence <- str_flatten(words, ___)

sentence
```


### Working with Unicode Characters

Handle strings with special characters or different encodings.

```{r, echo=TRUE}
# String with Unicode characters
unicode_text <- "Hello, \u4E16\u754C" # "Hello, World" in Mandarin

unicode_text
```

### Exercise: Unicode Strings

Create a string with Unicode characters:

```{r unicode, exercise = TRUE}
# Create a Unicode string
text <- "\u0048\u0065\u006C\u006C\u006F" # "Hello"

text
```


### Locale-Sensitive Operations
Sort or transform strings based on locale-specific rules.


```{r, echo=TRUE}
# Sort strings in a specific locale
library(stringi)
words <- c("ångstrom", "apple", "Ångström")
sorted <- stri_sort(words, locale = "sv") # Swedish locale

sorted
```

### Locale-Sensitive Sorting

Sort the following words using a specified locale:

```{r locale, exercise = TRUE}
words <- c("zebra", "apple", "banana")
sorted <- stri_sort(words, ___)

sorted
```

```{r quiz3}
quiz(
  question("Which function is used to extract substrings from a string?",
    answer("str_flatten()"),
    answer("str_split()"),
    answer("str_sub()", correct = TRUE)
  ),
  question("What does the `str_replace_all()` function do?",
    answer("Splits a string into parts."),
    answer("Joins a list of strings into one."),
    answer("Replaces all occurrences of a pattern in a string.", correct = TRUE)
  )
)
```

  
## Regular expressions

### What are Regular Expressions?
Regular expressions are sequences of characters that define a search pattern. They are widely used for string matching and manipulation.

### Key Functions
- `str_view()` and `str_view_all()`: Visualize regex matches in strings.

### Example
```{r, echo=TRUE}
library(stringr)
strings <- c("apple", "banana", "pear")
str_view(strings, "a")
```

### Find a Pattern
Use `str_view()` to visualize all occurrences of the letter "p" in the following strings:

```{r strview, exercise = TRUE}
strings <- c("apple", "banana", "pear")
str_view(strings, ___)
```


### Basic Regex Patterns



### Literal Characters
Match exact characters in strings.

```{r, echo=TRUE}
str_view(c("cat", "bat", "rat"), "cat")
```

### Metacharacters
Characters with special meanings, such as `.`, `*`, `+`, `?`, and `[]`.

```{r, echo=TRUE}
str_view(c("cat", "bat", "rat"), "c.t")
```

### Match Patterns
Write a pattern to match any three-letter word ending in "at":

```{r dot, exercise = TRUE}
words <- c("cat", "bat", "rat", "hat", "mat")
str_view(words, ___)
```

```{r dot-solution}
words <- c("cat", "bat", "rat", "hat", "mat")
str_view(words, ".at")
```


### Anchors and Character Classes

### Anchors
- `^`: Matches the start of a string.
- `$`: Matches the end of a string.

```{r, echo=TRUE}
str_view(c("apple", "banana", "pear"), "^a")
```

### Character Classes
- `[abc]`: Matches any character in the set.
- `[^abc]`: Matches any character not in the set.

### Shorthand Classes
- `\d`: Digits
- `\w`: Word characters
- `\s`: Whitespace


```{r, echo=TRUE}
str_view(c("123", "abc", "   "), "\\d")
# We need to use \\ to escape the backslash
```

### Use Anchors and Classes
Write a pattern to match strings that start with "a" and contain digits:

```{r matcha, exercise = TRUE}
strings <- c("apple1", "banana2", "pear3")
str_view(strings, ___)
```


### Quantifiers and Grouping

### Quantifiers
- `*`: 0 or more
- `+`: 1 or more
- `?`: 0 or 1
- `{n}`: Exactly `n` occurrences

```{r, echo=TRUE}
str_view(c("aa", "aaa", "aaaa"), "a{2,3}")
```

### Grouping and Capturing
Use `()` to group patterns and capture parts of a match.

```{r, echo=TRUE}
str_view(c("12-34", "56-78", "90-12"), "(\\d{2})-(\\d{2})")
```

### Quantifiers and Groups
Write a pattern to match three-letter words starting with "b" and ending with "g":

```{r matchbg, exercise = TRUE}
words <- c("bag", "big", "bug", "bog")
str_view(words, ___)
```


### Advanced Pattern Matching and Replacement

### Escaping Metacharacters
Use `\\` to match metacharacters literally.

```{r, echo=TRUE}
str_view(c("1+1=2", "2*2=4"), "\\+")
```

### Pattern Replacement
- `str_replace()`: Replace the first match.
- `str_replace_all()`: Replace all matches.

```{r, echo=TRUE}
str_replace("The cat sat on the mat.", "cat", "dog")
```

### Regex Flags
Modify regex behavior with flags (e.g., case-insensitive matching).

### Pattern Replacement
Replace all instances of "is" with "was" in the following string:

```{r iswas, exercise = TRUE}
sentence <- "This is a test. It is fun. Or, at least my teacher think it is. Weido!"
updated_sentence <- str_replace_all(sentence, ___, ___)

updated_sentence
```


### Quiz

```{r quiz4}
quiz(
  question("What does the `^` anchor match in a string?",
    answer("The start of a string.", correct = TRUE),
    answer("Any character in a set."),
    answer("0 or more occurrences.")
  ),
  question("Which metacharacter matches 0 or 1 occurrences of a pattern?",
    answer("*"),
    answer("?"),
    answer("+", correct = TRUE)
  ),
  question("What does the `\\d` shorthand class match?",
    answer("Digits", correct = TRUE),
    answer("Words"),
    answer("Whitespace")
  ),
  question("Which regex pattern matches strings that start with 'a' and end with 'e'?",
    answer("^a.e$"),
    answer("a*e"),
    answer("^ae$", correct = TRUE)
  ),
  question("What does the `str_replace_all()` function do?",
    answer("Replaces the first match."),
    answer("Replaces all matches.", correct = TRUE),
    answer("Splits a string into parts.")
  )
)
```


## Factors

Factors are used to represent categorical data in R. They are stored as integers with a corresponding set of character values.
In this part of the tutorial, based on [chapter 16 of R for data science](https://r4ds.hadley.nz/factors) you will learn what factors are, why they are useful, and how to work with them using the `forcats` package.
Factors can be unordered, for example, storing categories like gender, or ordered, for example, storing age ranges.

### Create a Factor
Create a factor to represent the following data: `"apple", "banana", "apple", "cherry"`

```{r createf, exercise = TRUE}
library(forcats)
fruits <- factor(c(___))

fruits
```


### Levels
Levels are the distinct categories in a factor. You can view and modify levels using `levels()`.

```{r, echo=TRUE}
# View and modify levels
levels(colors) <- c("red", "blue", "green")

colors
```

### Change Levels
Modify the levels of the `fruits` factor created earlier to capitalize the first letter of each fruit:

```{r levels, exercise = TRUE}
# Modify levels
levels(fruits) <- c(___)

fruits
```


### Ordered Factors

Sometimes the order of levels is important. You can create ordered factors using `factor(..., ordered = TRUE)`.

```{r, echo=TRUE}
sizes <- factor(c("small", "medium", "large", "medium"), 
                levels = c("small", "medium", "large"), 
                ordered = TRUE)

sizes
```

### Create an Ordered Factor
Create an ordered factor for the days of the week, with `"Monday"` as the first level and `"Sunday"` as the last:

```{r ordered, exercise = TRUE}
# Create ordered factor
days <- factor(c("Monday", "Tuesday", "Wednesday", "Sunday"), 
               levels = c(___), 
               ordered = TRUE)
days
```

### Manipulating Factors with forcats

Use `fct_reorder()` to reorder levels based on a variable.

```{r, echo=TRUE}
# Reorder levels based on frequency
fruits <- factor(c("apple", "banana", "apple", "cherry"))
fruits <- fct_infreq(fruits)
fruits
```

### Reorder Factors by Length
Reorder the `fruits` factor so that the levels are ordered by the length of the fruit name:

```{r reorder, exercise = TRUE}
# Reorder by length
fruits <- fct_reorder(fruits, ___)

fruits
```

### Combining and Collapsing Levels

Use `fct_collapse()` to combine multiple levels into one.

```{r, echo=TRUE}
# Combine levels
fruits <- fct_collapse(fruits, 
                        citrus = c("orange", "lemon"), 
                        tropical = c("banana", "mango"))
fruits
```

### Combine Levels

Combine the levels of the `fruits` factor into `"apple"` and `"other"`:

```{r combine, exercise = TRUE}
# Combine levels
fruits <- fct_collapse(fruits, 
                        apple = "apple", 
                        other = c(___))

fruits
```

### Quiz

```{r quiz5}
quiz(
  question("What is the purpose of factors in R?",
    answer("To represent categorical data.", correct = TRUE),
    answer("To store numerical data."),
    answer("To visualize data.")
  ),
  question("How do you create an ordered factor in R?",
    answer("factor(..., ordered = TRUE)", correct = TRUE),
    answer("factor(..., ordered = FALSE)"),
    answer("fct_reorder()")
  ),
  question("What does the `fct_collapse()` function do?",
    answer("Reorders factor levels."),
    answer("Combines multiple levels into one.", correct = TRUE),
    answer("Changes the order of factor levels.")
  )
)

```


## Dates and times

### Basics of Dates and Times
In this part of the tutorial, based on [chapter 16 of R for data science](https://r4ds.hadley.nz/dates-and-times.html) you will learn how to work with dates and times in R using the `lubridate` package.
Dates and times are represented as special objects in R.

### Key Functions
- `ymd()`, `mdy()`, `dmy()`: Parse dates.
- `now()`, `today()`: Get the current date and time.

```{r, echo=TRUE}
# Parse a date
library(lubridate)
date <- ymd("2023-12-31")
date

# Get the current date and time
current_time <- now()
print(current_time)
```

### Parse a Date
Parse the string `"31-12-2023"` as a date:

```{r parsedate, exercise = TRUE}
# Parse the date
date <- dmy("___")

date
```

### Extracting Parts of a Date/Time
You can extract specific components of a date or time using functions like `year()`, `month()`, `day()`, `hour()`, `minute()`, and `second()`.

```{r, echo=TRUE}
# Extract components
time <- now()
year <- year(time)
month <- month(time)
day <- day(time)

print(year)
print(month)
print(day)
```


### Extract Components
Extract the month and day from the following date:

```{r extractdate, exercise = TRUE}
date <- ymd("2023-12-31")
month <- month(date)
day <- day(date)
print(month)
print(day)
```

### Date Arithmetic

You can perform arithmetic on dates and times using `+` and `-` operators.

```{r, echo=TRUE}
# Add and subtract time
date <- ymd("2023-12-31")
next_week <- date + weeks(1)
last_month <- date - months(1)

next_week
last_month
```

### Adjust Dates
Find the date two weeks after `"2023-12-31"`:

```{r twoweeks, exercise = TRUE}
# Adjust the date
date <- ymd("2023-12-31")
adjusted_date <- date + weeks(___)
adjusted_date
```

### Time Zones

You can set or change the time zone of a datetime object using `with_tz()` or `force_tz()`.

```{r echo=TRUE}
# Set a time zone
time <- now()
time_utc <- with_tz(time, "UTC")
time_utc

# Force a time zone
time_forced <- force_tz(time, "America/New_York")
time_forced
```

### Change Time Zone
Convert the current time to UTC:

```{r tz, exercise = TRUE}
time <- now()
time_utc <- with_tz(time, ___)
time_utc
```


### Parsing and Formatting Dates
Use `parse_date_time()` to parse dates and `format()` to format them into strings.

```{r, echo=TRUE}
parsed_date <- parse_date_time("31-12-2023 23:59", "dmy HM")
formatted_date <- format(parsed_date, "%B %d, %Y")

parsed_date
formatted_date
```

### Format a Date
Format the date `"2023-12-31"` as `"December 31, 2023"`:

```{r formatdate, exercise = TRUE}
date <- ymd("2023-12-31")
formatted_date <- format(date, ___)
formatted_date
```


### Quiz

```{r quiz6}
quiz(
  question("What is the purpose of the `lubridate` package in R?",
    answer("To work with dates and times.", correct = TRUE),
    answer("To visualize data."),
    answer("To manipulate strings.")
  ),
  question("How do you extract the month from a date object in R?",
    answer("month()", correct = TRUE),
    answer("day()"),
    answer("year()")
  ),
  question("What does the `weeks()` function do in R?",
    answer("Adds a week to a date.", correct = TRUE),
    answer("Extracts the week of the year."),
    answer("Converts a date to a week.")
  ),
  question("What does the `with_tz()` function do?",
    answer("Forces a time zone on a date."),
    answer("Sets the time zone of a datetime object.", correct = TRUE),
    answer("Formats a date.")
  ),
  question("What does the `format()` function do?",
    answer("Parses dates and times."),
    answer("Formats dates and times into strings.", correct = TRUE),
    answer("Extracts components of a date.")
  ),
  question("What is the output of the following code? `parse_date_time('31-12-2023 23:59', 'dmy HM')`",
    answer("2023-12-31 23:59:00"),
    answer("December 31, 2023 11:59 PM"),
    answer("2023-12-31 11:59 PM", correct = TRUE)
  ),
  question('Which function parses dates in the format `"YYYY-MM-DD"`?',
    answer("ymd()", correct = TRUE),
    answer("mdy()"),
    answer("dmy()")
  )
)
```

## Missing value

In R, missing values are represented as `NA`. They can indicate missing or undefined data.

### Key Functions
- `is.na()`: Check for missing values.
- `anyNA()`: Check if there are any missing values in a dataset.

```{r, echo=TRUE}
data <- c(1, 2, NA, 4, NA, 6)
# Identify missing values
missing <- is.na(data)
missing

# Check if any values are missing
any_missing <- anyNA(data)
any_missing
```

### Identify Missing Values
Identify the missing values in the following vector:

```{r missing, exercise = TRUE}
data <- c(3, NA, 7, NA, 10)

missing <- is.na(___)
missing
```

### Removing Missing Values
You can remove missing values using `na.omit()` or by subsetting.

```{r, echo=TRUE}
data <- c(1, 2, NA, 4, NA, 6)
# Remove missing values
cleaned_data <- na.omit(data)
cleaned_data

# Remove missing values using subsetting
cleaned_data2 <- data[!is.na(data)]
cleaned_data2
```

### Remove Missing Values
Remove the missing values from the following vector:

```{r removemissing, exercise = TRUE}
data <- c(3, NA, 7, NA, 10)

cleaned_data <- ___

cleaned_data
```


### Replacing Missing Values
You can replace missing values with specific values using `ifelse()` or `replace()`.

```{r, echo=TRUE}
data <- c(1, 2, NA, 4, NA, 6)

data_filled <- ifelse(is.na(data), 0, data)
data_filled
```

### Replace Missing Values
Replace the missing values in the following vector with the value `-1`:

```{r replaceMissing, exercise = TRUE}
data <- c(3, NA, 7, NA, 10)
data_filled <- ifelse(___, ___, data)
data_filled
```

### Summarizing Data with Missing Values

Many functions in R can handle missing values using the `na.rm = TRUE` argument.


```{r, echo=TRUE}
data <- c(1, 2, NA, 4, NA, 6)
# Calculate the sum, ignoring missing values
sum_data <- sum(data, na.rm = TRUE)
sum_data

# Calculate the mean, ignoring missing values
mean_data <- mean(data, na.rm = TRUE)
mean_data
```

### Summarize Data
Calculate the mean of the following vector, ignoring missing values:

```{r summaryMissing, exercise = TRUE}
data <- c(3, NA, 7, NA, 10)

mean_data <- mean(___)
mean_data
```


### Missing Values in Data Frames
You can handle missing values in data frames using similar techniques.

```{r, echo=TRUE}
data <- data.frame(
  x = c(1, 2, NA, 4),
  y = c(NA, 2, 3, 4)
)

cleaned_data <- na.omit(data)
cleaned_data

# Replace missing values with column means
data$x[is.na(data$x)] <- mean(data$x, na.rm = TRUE)
data$y[is.na(data$y)] <- mean(data$y, na.rm = TRUE)
data
```

### Handle Missing Values
Replace the missing values in the following data frame with 0:

```{r handleMissing, exercise = TRUE}
data <- data.frame(
  x = c(1, 2, NA, 4),
  y = c(NA, 2, 3, 4)
)

data[is.na(data)] <- ___
data
```


As we have seen, R codes missing values as `NA`. But it also has other special values for certain situations, like `Inf` for infinity and `-Inf` for negative infinity. It also has codes for not-a-number: `NaN` which is used to represent undefined or unrepresentable values. It also has `NA_integer_`, `NA_real_`, `NA_complex_`, and `NA_character_` for missing values of different types. Finally, it has `NULL` which is used to represent the absence of a value.

Think carefully about these when you are using/writing functions as they might not all evaluate to the same result.

### Quiz

```{r quiz7}
quiz(
  question("What is the purpose of the `is.na()` function in R?",
    answer("To check for missing values.", correct = TRUE),
    answer("To calculate the mean of a vector."),
    answer("To replace missing values.")
  ),
  question("What does the `na.rm = TRUE` argument do in R?",
    answer("Removes missing values from a calculation.", correct = TRUE),
    answer("Replaces missing values with the mean."),
    answer("Checks if any values are missing.")
  ),
  question("What is the output of the following code? `sum(c(1, 2, NA, 4), na.rm = TRUE)`",
    answer("7", correct = TRUE),
    answer("NA"),
    answer("3")
  ),
  question("What is the special value used to represent infinity in R?",
    answer("NA"),
    answer("Inf", correct = TRUE),
    answer("NaN")
  ),
  question("What is the special value used to represent not-a-number in R?",
    answer("NA"),
    answer("Inf"),
    answer("NaN", correct = TRUE)
  )
)
```

## Packages

Much of R strength rely on the ecosystem of packages that it comes with.

In many languages, library is the term used to describe what an R package is (confusingly, you load a package in memory using the function `library()`).

What is a package? 

It is a collection of functions that allow you to perform specific types of operations, usually because they are not available among the functions the software comes prepackaged with.

R has packages for tasks like data loading, data wrangling, visualisation, modelling and many more tasks.

## Installing packages

Packages are installed with the function `install.packages()` and updated with `update.packages()`

```{r, eval = FALSE, echo=TRUE}
install.packages("name of the package")
install.packages(c("package_1", "package_2"))

```

Once you have installed a package, you need to load it before you can use the functions in it.

```{r, eval = FALSE, echo=TRUE}
library(name_of_the_package)

# What most people do (me included) 
function_from_package(args)

# What you should do in your scripts
# This is safe if two packages have a function with the same name
name_of_the_package::function_from_package(args)
```


### Congratulations

Congratulations! You have learned a lot in this tutorial! Well done for sticking with it until the end.


