---
title: "Probability in R"
output: 
  learnr::tutorial:
    df_print: paged
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Learn about probability calculations in R through interactive exercises"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis::gradethis_setup()

library(NHANES)
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.pipe = "|>")
```

# Probability in R

In this chapter we will go over probability computations in R.

## Basic probability calculations

Let's create a vector of outcomes from one to 6, using the `seq()` function to create such a sequence:

```{r outcomes-setup}
# Create the initial outcomes vector that will be used in subsequent exercises
outcomes <- seq(1, 6)
```

```{r outcomes, exercise=TRUE, exercise.setup="outcomes-setup"}
# Create a vector from 1 to 6 using seq()
# Assign it to the variable outcomes and show the result

outcomes <- _______

outcomes
```

::: {#outcomes-hint}
Remember: you can use ?seq to get help on the seq() function
:::

```{r outcomes-solution}
outcomes <- seq(1, 6)

outcomes
```

```{r outcomes-check}
grade_this_code()
```

### Logical values

Now let's create a vector of logical values based on whether the outcome in each position is equal to 1. Remember that `==` tests for equality of each element in a vector:

```{r outcomes-bool-setup}
# Make outcomes available
outcomes <- seq(1, 6)
```

```{r outcomes-bool, exercise=TRUE, exercise.setup="outcomes-bool-setup"}
# Create a vector of logical values based on whether 
# the outcome in each position is equal to 1.
# Assign it to the variable outcome_1_is_true and show the result

outcome_1_is_true <- _________

outcome_1_is_true
```

```{r outcomes-bool-hint}
# You can use the == operator to test for equality
# Let's see what happens when we test for equality of the outcomes vector with 1

outcomes == 1
```

```{r outcomes-bool-solution}
outcome_1_is_true <- outcomes == 1

outcome_1_is_true
```

```{r outcomes-bool-check}
grade_this_code()
```

Remember that the simple probability of an outcome is number of occurrences of the outcome divided by the total number of events. To compute a probability, we can take advantage of the fact that TRUE/FALSE are equivalent to 1/0 in R. The formula for the mean (sum of values divided by the number of values) is thus exactly the same as the formula for the simple probability! So, we can compute the probability of the event by simply taking the mean of the logical vector.

Use the `mean()` function to calculate the probability of the event that the outcome is 1.

```{r probability-setup}
# Make outcomes and outcome_one_is_true available
outcomes <- seq(1, 6)
outcome_1_is_true <- outcomes == 1
```

```{r probability, exercise=TRUE, exercise.setup="probability-setup"}
# Assign the result to the variable p_one_is_true and show the result
p_1_is_true <- ___________

p_1_is_true
```

```{r probability-solution}
p_1_is_true <- mean(outcome_1_is_true)

p_1_is_true
```

```{r probability-check}
grade_this_code()
```

## Empirical Frequency

Now let's explore how to calculate empirical probabilities using real data. We'll analyze rainfall patterns in San Francisco to understand the probability of rain on any given day.

We'll use a dataset containing daily precipitation records for San Francisco. First, let's load and prepare the data by removing unnecessary columns:

```{r prepare-sfrain, message=FALSE, exercise=TRUE}
# we will remove the STATION and NAME variables
# since they are identical for all rows

sf_rain <- read_csv("data/SanFranciscoRain-1329219.csv") |>
  dplyr::select(-STATION, -NAME)
```

### Examine the data

Let's examine our data using the `glimpse()` function. This function provides a concise summary of each variable in our dataset:

```{r view-sfrain, exercise=TRUE, exercise.setup="prepare-sfrain"}
glimpse(sf_rain)
```


### Calculate P(Rain in SF)
The data frame contains a variable called `PRCP` (precipitation) which measures the amount of rainfall in inches for each day. To calculate the probability of rain, we first need to create a binary variable that indicates whether it rained on each day (1) or not (0).

Within a pipe, we use the `mutate()` function to create a new variable. Here, we'll create a new variable called `rain_today` and use it to indicate whether it rained on each day (1) or not (0). This will return a data frame with the new variable added as a new column.

The expression `PRCP > 0` creates a logical (TRUE/FALSE) value, which we convert to 1/0 using `as.integer()`:

```{r rain-today, exercise=TRUE, exercise.setup="view-sfrain"}
sf_rain <-
  sf_rain |>
    mutate(rain_today = as.integer(PRCP > 0))

glimpse(sf_rain)
```

Now we can calculate the probability of rain by taking the mean of the `rain_today` variable.

```{r probability-rain-in-sf, exercise=TRUE, exercise.setup="rain-today"}
# Calculate the probability of rain by taking the mean of rain_today
# Assign the result to p_rain_in_sf and show the result

p_rain_in_sf <-
  __________

p_rain_in_sf
```

```{r probability-rain-in-sf-hint}
# Remember: we access a variable by using the dollar sign: 
sf_rain$rain_today
```

```{r probability-rain-in-sf-solution}
p_rain_in_sf <- mean(sf_rain$rain_today)

p_rain_in_sf
```

```{r probability-rain-in-sf-check}
grade_this_code()
```

### Bonus: Taking advantage of the pipe

A cleaner approach might be to use the pipe operator to chain together the `mutate()` and `mean()` functions.

Think of this as a processing pipeline to transform the data as needed before doing the final calculation we want, rather than continually reassigning to new variables. This approach can be more concise and easier to read.

```{r prain-pipe, exercise=TRUE, exercise.setup="rain-today"}
# Use the pipe operator to chain together the logical expression and the mean function
p_rain_in_sf <-
  _________

p_rain_in_sf
```

```{r prain-pipe-hint-1}
# We use the `summarise()` function to summarise the 
# observations in the data frame. 
# 
# Pass `mean(rain_today)` as the function to 
# `summarise()` the data frame.
```

```{r prain-pipe-hint-2}
# We can make the output even cleaner by pulling the mean from the 
# data frame by adding `pull()` to the end of the pipeline.

# The structure should be:
  . . . |>
  summarise(mean(rain_today)) |>
  pull()
```

```{r prain-pipe-solution}
# Default ----
p_rain_in_sf <-
  sf_rain |>
  mutate(rain_today = as.integer(PRCP > 0)) |>
  summarise(mean(rain_today))

p_rain_in_sf

# Cleaner: pull the mean from the data frame ----
p_rain_in_sf <-
  sf_rain |>
  mutate(rain_today = as.integer(PRCP > 0)) |>
  summarise(mean(rain_today)) |>
  pull()

p_rain_in_sf
```

```{r prain-pipe-check}
grade_this_code()
```

## Joint Probability

The joint probability is the probability of two events occurring together. In our case, we want to find the probability that someone is both over 70 years old AND unhealthy.

### Set up the data and logical variables

We'll explore this using health data from the NHANES (National Health and Nutrition Examination Survey) dataset. First, let's prepare our data by creating two binary variables:

1.  `over70`: TRUE if someone is over 70 years old
2.  `unhealthy`: TRUE if they reported any days of poor physical health (`DaysPhysHlthBad > 0`)

Here's how we create these variables:

```{r health-setup, exercise=TRUE}
# Create a data frame with Over70 and Unhealthy variables
health_df <-
  NHANES |>
    mutate(
      over70 = Age > 70,
      unhealthy = DaysPhysHlthBad > 0
    ) |>
    dplyr::select(unhealthy, over70) |>
    drop_na()

glimpse(health_df)
```

### Calculate the separate probabilities:

Let's calculate the probability of being over 70 years old in our dataset:

```{r over70-prob, exercise=TRUE, exercise.setup="health-setup"}
# Calculate the probability of being over 70
# Use summarise() to calculate the mean of over70
# Assign the result to p_over70 and show the result

p_over70 <-
  health_df |>
    ___________

p_over70
```

::: {#over70-prob-hint}
**The structure should be:**

```r
new_name <- 
  df |>
    summarise(new_name = mean(...))
``` 

Or, using pull() which extracts the calculated value from the dataframe:

```r
new_name <- 
  df |>
    summarise(mean(...)) |>
    pull()
```
:::

```{r over70-prob-solution}
# Assign to a new variable ----
p_over70 <-
  health_df |>
    summarise(p_over70 = mean(over70))

p_over70

# Or pull the mean from the data frame ----
p_over70 <-
  health_df |>
    summarise(mean(over70)) |>
    pull()

p_over70
```

```{r over70-prob-check}
grade_this({
  # Check if they used summarize and mean
  if (!grepl("summarise.*mean\\(.*over70.*\\)", .user_code)) {
    fail(
      "Make sure to use summarise() and mean(over70) to calculate the probability"
    )
  }
  if (typeof(.result) == "list") {
    .result <- pull(.result)
  }
  # Check if result matches solution
  pass_if_equal()
  fail()
})
```

Now let's calculate the probability of being unhealthy:

```{r unhealthy-prob, exercise=TRUE, exercise.setup="health-setup"}
# Calculate the probability of being unhealthy
# Use summarise() to calculate the mean of unhealthy
# Assign the result to p_unhealthy and show the result

p_unhealthy <-
  health_df |>
    ___________

p_unhealthy
```

::: {#unhealthy-prob-hint}
**The structure should be:**

```r
new_name <- 
  df |>
    summarise(new_name = mean(...))
``` 

Or, using pull() which extracts the calculated value from the dataframe:

```r
new_name <- 
  df |>
    summarise(mean(...)) |>
    pull()
```
:::

```{r unhealthy-prob-solution}
# Assign to a new variable ----
p_unhealthy <-
  health_df |>
    summarise(p_unhealthy = mean(unhealthy))

p_unhealthy

# Or pull the mean from the data frame ----
p_unhealthy <-
  health_df |>
    summarise(mean(unhealthy)) |>
    pull()

p_unhealthy
```

```{r unhealthy-prob-check}
grade_this({
  # Check if they used summarize and mean
  if (!grepl("summarise.*mean\\(.*unhealthy.*\\)", .user_code)) {
    fail(
      "Make sure to use summarise() and mean(unhealthy) to calculate the probability"
    )
  }
  if (typeof(.result) == "list") {
    .result <- pull(.result)
  }
  # Check if result matches solution
  pass_if_equal()
  fail()
})
```

### Calculate the joint probability

To calculate this, we need to:

1.  Create a new variable that is TRUE (1) only when both conditions (`over70` and `unhealthy`) are TRUE
2.  Calculate the mean of this new variable to get the probability

We can use multiplication (\*) with logical values because:

-   TRUE \* TRUE = 1
-   TRUE \* FALSE = 0
-   FALSE \* FALSE = 0

Try it yourself:

```{r joint-prob, exercise=TRUE, exercise.setup="health-setup"}
# Calculate the probability of being both over 70 AND unhealthy
# First create a new variable that multiplies over70 and Unhealthy
# Then calculate the mean of this new variable
# Assign the result to p_both and show the result

p_both <-
  health_df |>
    mutate(
      both = ________
    ) |>
    summarise(p_both = mean(both))

p_both
```

```{r joint-prob-hint-1}
# When we multiply two logical values, 
# the result is TRUE (1) only if both are TRUE
both = over70 * unhealthy
```

```{r joint-prob-solution}
p_both <-
  health_df |>
    mutate(
      both = over70 * unhealthy
    ) |>
    summarise(mean(both)) |>
  pull()

p_both
```

```{r joint-prob-check}
grade_this({
  # Check if they used both Over70 and Unhealthy
  if (!grepl("both.*=.*over70.*\\*.*unhealthy", .user_code) && !grepl("both.*=.*unhealthy*\\*.*over70", .user_code)) {
    fail(
      "Make sure to multiply over70 and unhealthy to get the joint probability"
    )
  }
  if (typeof(.result) == "list") {
    .result <- pull(.result)
  }
  # Check if result matches solution
  pass_if_equal()
  fail()
})
```

## Conditional Probability

Conditional probability is the probability of one event occurring, given that another event has occurred. For example, we might want to know: what's the probability of being unhealthy, specifically among people who are over 70 years old?

### Calculate P(Unhealthy | Over 70)

Now we'll calculate P(Unhealthy \| Over 70) - the probability of being unhealthy given that someone is over 70.

To calculate this:

1.  First, we need to focus only on people over 70 (using `filter()`)
2.  Then, within this subset, calculate the probability of being unhealthy

This is different from the joint probability because we're only looking at the subset of people who are over 70, rather than the entire population. Try it:

```{r conditional-prob, exercise=TRUE, exercise.setup="health-setup"}
# Calculate P(Unhealthy | Over 70)
# First filter to only include people over 70
# Then calculate the probability of being unhealthy
# Assign the result to p_unhealthy_given_over70 and show the result

p_unhealthy_given_over70 <-
  health_df |>
    filter(________) |>
    summarise(p_unhealthy = mean(unhealthy))

p_unhealthy_given_over70
```

```{r conditional-prob-hint-1}
# Use filter() to keep only rows where over70 is TRUE
```

```{r conditional-prob-hint-2}
# The structure should be:
filter(over70 == TRUE)
```

```{r conditional-prob-solution}
p_unhealthy_given_over70 <-
  health_df |>
    filter(over70 == TRUE) |>
    summarise(p_unhealthy = mean(unhealthy)) |>
  pull()

p_unhealthy_given_over70
```

```{r conditional-prob-check}
grade_this({
  # Check if they filtered for Over70
  if (!grepl("filter.*over70.*TRUE", .user_code)) {
    fail(
      "Make sure to filter for over70 == TRUE before calculating the probability"
    )
  }
  if (typeof(.result) == "list") {
    .result <- pull(.result)
  }
  # Check if result matches solution
  pass_if_equal()
  fail()
})
```

### Reverse Conditional Probability P(Over 70 | Unhealthy)

Finally, let's look at the reverse question: among unhealthy people, what proportion are over 70? This is written as P(Over70 \| Unhealthy) - the probability of being over 70 given that someone is unhealthy.

Notice how this is different from P(Unhealthy \| Over70) that we calculated earlier:

-   P(Unhealthy \| Over70) asks: "Among people over 70, what proportion are unhealthy?"
-   P(Over70 \| Unhealthy) asks: "Among unhealthy people, what proportion are over 70?"

Calculate this reverse conditional probability:

```{r reverse-conditional, exercise=TRUE, exercise.setup="health-setup"}
# Calculate P(Over70 | Unhealthy)
# First filter to only include unhealthy people
# Then calculate the probability of being over 70
# Assign the result to p_over70_given_unhealthy and show the result

p_over70_given_unhealthy <-
  health_df |>
    ________

p_over70_given_unhealthy
```

```{r reverse-conditional-hint-1}
# First filter for Unhealthy == TRUE
# Then calculate mean(over70)
```

```{r reverse-conditional-hint-2}
# The structure should be:
filter(unhealthy == TRUE) |>
  summarise(p_over70 = mean(over70))
```

```{r reverse-conditional-solution}
p_over70_given_unhealthy <-
  health_df |>
    filter(unhealthy == TRUE) |>
    summarise(p_over70 = mean(over70)) |>
  pull()

p_over70_given_unhealthy
```

```{r reverse-conditional-check}
grade_this({
  # Check if they filtered for Unhealthy
  if (!grepl("filter.*unhealthy.*TRUE", .user_code)) {
    fail(
      "Make sure to filter for unhealthy == TRUE before calculating the probability"
    )
  }
  if (typeof(.result) == "list") {
    .result <- pull(.result)
  }
  # Check if result matches solution
  pass_if_equal()
  fail()
})
```
